Efficient Calculation of Improved Simultaneous Testing Bands for QQplots with \proglang{R} Package \pkg{qqconf}

### 2021-01-24 article/manuscript.tex
adding article

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE}
<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
@
\raggedbottom
\graphicspath{ {./images/} }
\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENT}{\# #1}}
\makeatletter
\newcommand{\distas}[1]{\mathbin{\overset{#1}{\kern\z@\sim}}}%
\newsavebox{\mybox}\newsavebox{\mysim}
\newcommand{\distras}[1]{%
  \savebox{\mybox}{\hbox{\kern3pt$\scriptstyle#1$\kern3pt}}%
  \savebox{\mysim}{\hbox{$\sim$}}%
  \mathbin{\overset{#1}{\kern\z@\resizebox{\wd\mybox}{\ht\mysim}{$\sim$}}}%
}
\makeatother

@@

More specifically, in each scenario, we randomly generate $n=1000$ $2 \times 2$ tables under the null hypothesis, where each table contains $s$ observations, with $s=200$ in scenario 1 and $s=20$ in scenario 2.  For each table, the $s$ observations are i.i.d.\ with probability $q_{i,j}$ of falling in cell $(i,j)$, for $i=0,1, j=0,1$, where $q_{1,1} = a*b$, $q_{1,0} = a*(1-b)$, $q_{0,1} = (1-a)*b$, and $q_{0,0} = (1-a)*(1-b)$, with a=.15 and b=.4.  For each table, let $X_{i,j}$ denote the observed count in cell $(i,j)$.  If any table has $X_{0,0}+X_{0,1} = 0$ or $= s$, we discard the table and draw a new one, because that would imply that one of the rows of the table is empty, in which case the Pearson Chi-Squared test statistic is not defined.  Similarly, if any table has $X_{0,0}+X_{1,0} = 0$ or $=s$, we discard the table and draw a new one. The code to do this is shown below:
%
<<contingency_tables>>=
set.seed(8675309)
# Generate contingency tables for variables as defined above
generate_contingency_tables <- function(n, s, a, b) {

  cell_probs <- c(a * b, a * (1 - b), b * (1 - a), (1 - b) * (1 - a))
  num_sims <- 0
  sim_mat <- matrix(NA, nrow = n, ncol = 4)

  # Generate n simulations and append the contingency tables to a matrix
  while (num_sims < n) {

    proposed_sim <- t(rmultinom(n = 1, size = s, prob = cell_probs))

    # Check to make sure that the chi-square statistic will be defined
    if (proposed_sim[, 1] + proposed_sim[, 2] != 0 && 
        proposed_sim[, 1] + proposed_sim[, 3] != 0 &&
        proposed_sim[, 3] + proposed_sim[, 4] != 0 &&
        proposed_sim[, 4] + proposed_sim[, 2]) {

      num_sims <- num_sims + 1
      sim_mat[num_sims, ] <- proposed_sim

    }

  }

  return(sim_mat)

}

scenario_1 <- generate_contingency_tables(n = 1000, s = 200, a = .15, b = .4)
scenario_2 <- generate_contingency_tables(n = 1000, s = 20, a = .15, b = .4)
@
%

@@

%
<<p_values>>=
# Compute Pearson Chi-Square Test p-values for Set of Contingency Tables
get_p_values_from_c_tables <- function(c_tables, s) {

  chisq_numerator <- (
    (c_tables[, 1] * c_tables[, 4] - c_tables[, 2] * c_tables[, 3]) / s
  ) ^ 2
  chisq_denominator <- (c_tables[, 1] + c_tables[, 2]) * 
    (c_tables[, 1] + c_tables[, 3]) * 
    (c_tables[, 3] + c_tables[, 4]) * 
    (c_tables[, 2] + c_tables[, 4]) / 
    (s ^ 3)
  chisq_stats_vec <- chisq_numerator / chisq_denominator
  p_vals <- pchisq(q = chisq_stats_vec, df = 1, lower.tail = FALSE)
  return(p_vals)

}

pvals_scenario_1 <- get_p_values_from_c_tables(scenario_1, s = 200)
pvals_scenario_2 <- get_p_values_from_c_tables(scenario_2, s = 20)
@
%

@@

<<chisq_qq, fig=TRUE, include=FALSE>>=
library(qqconf)
bounds <- get_bounds_two_sided(alpha = .05, n = 1000)

# Plot 3: Q-Q plot of P-values from Chi-Squared Test
# for Independence in a 2 x 2 Table

plot(
  bounds$x, 
  sort(pvals_scenario_2), 
  type = "l", 
  col = "red", 
  xlab = "Expected Uniform Quantile",
  ylab = "Observed P-value"
)

# Add the 45-degree line to the plot
lines(bounds$x, bounds$x)

# Add the lower bound line for the level .05 
# simultaneous testing bands from equal local levels.
lines(bounds$x, bounds$lower_bound)

# Add the upper bound line for the level .05 
# simultaneous testing bands from equal local levels.
lines(bounds$x, bounds$upper_bound)

# Add the Q-Q plot for scenario 1.
lines(bounds$x, sort(pvals_scenario_1), col = "blue")

legend(
  x = .02, 
  y = .95, 
  legend = c("s=20","s=200"), 
  col = c("red","blue"), 
  lty = 1
)
@

@@

<<chisq_qq_log, fig=TRUE, include=FALSE>>=
# Plot 4: Q-Q plot, on the -log10 Scale, of P-values 
# from Chi-Squared Test for Independence in a 2 x 2 Table

plot(
  -log10(bounds$x), 
  -log10(sort(pvals_scenario_2)),
  type = "l",
  col = "red",
  xlab = "-log10(Expected Uniform Quantile)",
  ylab = "-log10(Observed P-value)"
)

# Add the 45-degree line to the plot
lines(-log10(bounds$x), -log10(bounds$x))

# Add the lower bound line for the level .05 
# simultaneous testing bands from equal local levels.
lines(-log10(bounds$x), -log10(bounds$lower_bound))

# Add the upper bound line for the level .05 
# simultaneous testing bands from equal local levels.
lines(-log10(bounds$x), -log10(bounds$upper_bound))

# Add the Q-Q plot for scenario 1, with axes on the -log10 scale.
lines(-log10(bounds$x), -log10(sort(pvals_scenario_1)), col = "blue")

legend(
  x = .2, 
  y = 3, 
  legend = c("s=20", "s=200"), 
  col = c("red","blue"), 
  lty = 1
)
@

>>>>

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}

\raggedbottom
\graphicspath{ {./images/} }
\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENT}{\# #1}}
\makeatletter
\newcommand{\distas}[1]{\mathbin{\overset{#1}{\kern\z@\sim}}}%
\newsavebox{\mybox}\newsavebox{\mysim}
\newcommand{\distras}[1]{%
  \savebox{\mybox}{\hbox{\kern3pt$\scriptstyle#1$\kern3pt}}%
  \savebox{\mysim}{\hbox{$\sim$}}%
  \mathbin{\overset{#1}{\kern\z@\resizebox{\wd\mybox}{\ht\mysim}{$\sim$}}}%
}
\makeatother

@@

More specifically, in each scenario, we randomly generate $n=1000$ $2 \times 2$ tables under the null hypothesis, where each table contains $s$ observations, with $s=200$ in scenario 1 and $s=20$ in scenario 2.  For each table, the $s$ observations are i.i.d.\ with probability $q_{i,j}$ of falling in cell $(i,j)$, for $i=0,1, j=0,1$, where $q_{1,1} = a*b$, $q_{1,0} = a*(1-b)$, $q_{0,1} = (1-a)*b$, and $q_{0,0} = (1-a)*(1-b)$, with a=.15 and b=.4.  For each table, let $X_{i,j}$ denote the observed count in cell $(i,j)$.  If any table has $X_{0,0}+X_{0,1} = 0$ or $= s$, we discard the table and draw a new one, because that would imply that one of the rows of the table is empty, in which case the Pearson Chi-Squared test statistic is not defined.  Similarly, if any table has $X_{0,0}+X_{1,0} = 0$ or $=s$, we discard the table and draw a new one. The code to do this is shown below:
%
\begin{Schunk}
\begin{Sinput}
R> set.seed(8675309)
R> # Generate contingency tables for variables as defined above
R> generate_contingency_tables <- function(n, s, a, b) {
+    
+    cell_probs <- c(a * b, a * (1 - b), b * (1 - a), (1 - b) * (1 - a))
+    num_sims <- 0
+    sim_mat <- matrix(NA, nrow = n, ncol = 4)
+    
+    # Generate n simulations and append the contingency tables to a matrix
+    while (num_sims < n) {
+      
+      proposed_sim <- t(rmultinom(n = 1, size = s, prob = cell_probs))
+      
+      # Check to make sure that the chi-square statistic will be defined
+      if (proposed_sim[, 1] + proposed_sim[, 2] != 0 && 
+          proposed_sim[, 1] + proposed_sim[, 3] != 0 &&
+          proposed_sim[, 3] + proposed_sim[, 4] != 0 &&
+          proposed_sim[, 4] + proposed_sim[, 2]) {
+        
+        num_sims <- num_sims + 1
+        sim_mat[num_sims, ] <- proposed_sim
+        
+      }
+      
+    }
+    
+    return(sim_mat)
+    
+  }
R> scenario_1 <- generate_contingency_tables(n = 1000, s = 200, a = .15, b = .4)
R> scenario_2 <- generate_contingency_tables(n = 1000, s = 20, a = .15, b = .4)
\end{Sinput}
\end{Schunk}
%

@@

%
\begin{Schunk}
\begin{Sinput}
R> # Compute Pearson Chi-Square Test p-values for Set of Contingency Tables
R> get_p_values_from_c_tables <- function(c_tables, s) {
+    
+    chisq_numerator <- (
+      (c_tables[, 1] * c_tables[, 4] - c_tables[, 2] * c_tables[, 3]) / s
+    ) ^ 2
+    chisq_denominator <- (c_tables[, 1] + c_tables[, 2]) * 
+      (c_tables[, 1] + c_tables[, 3]) * 
+      (c_tables[, 3] + c_tables[, 4]) * 
+      (c_tables[, 2] + c_tables[, 4]) / 
+      (s ^ 3)
+    chisq_stats_vec <- chisq_numerator / chisq_denominator
+    p_vals <- pchisq(q = chisq_stats_vec, df = 1, lower.tail = FALSE)
+    return(p_vals)
+    
+  }
R> pvals_scenario_1 <- get_p_values_from_c_tables(scenario_1, s = 200)
R> pvals_scenario_2 <- get_p_values_from_c_tables(scenario_2, s = 20)
\end{Sinput}
\end{Schunk}
%

@@

\begin{Schunk}
\begin{Sinput}
R> library(qqconf)
R> bounds <- get_bounds_two_sided(alpha = .05, n = 1000)
R> # Plot 3: Q-Q plot of P-values from Chi-Squared Test
R> # for Independence in a 2 x 2 Table
R> 
R> plot(
+    bounds$x, 
+    sort(pvals_scenario_2), 
+    type = "l", 
+    col = "red", 
+    xlab = "Expected Uniform Quantile",
+    ylab = "Observed P-value"
+  )
R> # Add the 45-degree line to the plot
R> lines(bounds$x, bounds$x)
R> # Add the lower bound line for the level .05 
R> # simultaneous testing bands from equal local levels.
R> lines(bounds$x, bounds$lower_bound)
R> # Add the upper bound line for the level .05 
R> # simultaneous testing bands from equal local levels.
R> lines(bounds$x, bounds$upper_bound)
R> # Add the Q-Q plot for scenario 1.
R> lines(bounds$x, sort(pvals_scenario_1), col = "blue")
R> legend(
+    x = .02, 
+    y = .95, 
+    legend = c("s=20","s=200"), 
+    col = c("red","blue"), 
+    lty = 1
+  )
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
R> # Plot 4: Q-Q plot, on the -log10 Scale, of P-values 
R> # from Chi-Squared Test for Independence in a 2 x 2 Table
R> 
R> plot(
+    -log10(bounds$x), 
+    -log10(sort(pvals_scenario_2)),
+    type = "l",
+    col = "red",
+    xlab = "-log10(Expected Uniform Quantile)",
+    ylab = "-log10(Observed P-value)"
+  )
R> # Add the 45-degree line to the plot
R> lines(-log10(bounds$x), -log10(bounds$x))
R> # Add the lower bound line for the level .05 
R> # simultaneous testing bands from equal local levels.
R> lines(-log10(bounds$x), -log10(bounds$lower_bound))
R> # Add the upper bound line for the level .05 
R> # simultaneous testing bands from equal local levels.
R> lines(-log10(bounds$x), -log10(bounds$upper_bound))
R> # Add the Q-Q plot for scenario 1, with axes on the -log10 scale.
R> lines(-log10(bounds$x), -log10(sort(pvals_scenario_1)), col = "blue")
R> legend(
+    x = .2, 
+    y = 3, 
+    legend = c("s=20", "s=200"), 
+    col = c("red","blue"), 
+    lty = 1
+  )
\end{Sinput}
\end{Schunk}
